Machine Learning can Advance Theory Formation in the Social Sciences

### 2023-07-11 presentation_isdp2023.qmd
update 2023-07-11 11:52:24.053328

---
title: "Machine Learning can Advance Theory Formation in the Social Sciences"
author: "Caspar J. Van Lissa<sup>1</sup>"
date: "<sup>1</sup>: Tilburg University, dept. Methodology & Statistics<br>Funded by NWO Veni grant  VI.Veni.191G.090"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
output: 
  ioslides_presentation:
    widescreen: true
    css: slide.css
bibliography      : ["manuscript/references.bib", "presentation/all_refs.bib"]
---

```{r setup, include=FALSE}
#css: https://maxcdn.bootstrapcdn.com/bootstrap/3.3.7/css/bootstrap.min.css
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

# Conceptual introduction

## Central thesis

@@

## Defining theory

@@

```{r, out.width="60%"}
library(tidySEM)
lay <- get_layout("the1", "the2",
                  "hyp1", "hyp2",
                  "pat1", "obs2",
                  "obs1", "con2", rows = 4)
nodes <- data.frame(
  name = as.vector(lay),
  label = c("Theory", "Working hypothesis", "Pattern", "Observation", "Theory", "Hypothesis", "Observation", "Test"),
  shape = rep("rect", length(lay)),
  group = rep(c("Inductive", "Deductive"), each = 4)
  )
edges <- data.frame(
  from = c("the1", "hyp1", "pat1", "the2", "hyp2", "obs2"),
  to = c("hyp1", "pat1", "obs1", "hyp2", "obs2", "con2"),
  arrow = rep(c("first", "last"), each = 3),
  group = rep(c("Inductive", "Deductive"), each = 3)
)
g <- prepare_graph(edges = edges, layout = lay, nodes = nodes)
plot(g)
```

* Lack of theory cannot be overcome by improving deductive research
* Theory formation requires inductive (exploratory) research [@creswellDesigningConductingMixed2017]

## Empirical cycle

```{r, out.width="40%"}
knitr::include_graphics("presentation/empiricalcycle.png")
```

De Groot’s empirical cycle: a model of knowledge production through scientific research [@degrootMethodologieGrondslagenVan1961]

@@

**NEW**: Implications for the theory crisis have not yet been discussed

## Machine learning vs unguided exploration

```{r, out.width="100%"}
knitr::include_graphics("Figure Machine learning 4.png")
```

# Advantages of Machine Learning

@@

## Holistic approach

@@

* Assumption for causal interpretation
    + Multicollinearity
* Good theory incorporates most important causes
    + Cannot assess the relative utility across studies
* Include potential predictors from various theories, and undertheorized factors

@@

## Person-centered approaches

@@

* E.g., latent class analyses
    + "Which individuals are similar?"
    + Other *unsupervised learning* methods exist
* RI-CLPM and DSEM
    + Rarely explain heterogeneity in within-person effects
    + Regularization
* Tree-based models
    + Group individuals based on predictors to maximize homogeneity of the outcome
    + "Why are these individuals similar?"

@@

When using data models to guide theory formation, generalizability is essential

@@

* cross-validation to estimate predictive accuracy

@@

```{r, out.width="40%"}
library(ggplot2)
colors <- c("Variance" = "red", "Bias" = "green", "Total error" = "black")

@@

## From models to theory

@@

* abductive formal theory construction (AFTC) framework
    + construct formal theory
    + Generate data
    + Mathematically compare to empirical data
    + like a posterior predictive check in Bayes
    + Discrepancies: amend formal theory

@@

# Example: Text Mining Systematic Review

## Emotion regulation in adolescence

Developmentally sensitive period [@zimmermannEmotionRegulationEarly2014]

20% develop psychopathology [@leeAdolescentMentalHealth2014]

Potentially lifelong implications for mental health and well-being

```{r, out.height="20%"}
knitr::include_graphics("trajectories.jpg")
```

<!-- ## Knowledge gap -->

<!-- Substantial empirical research, but: -->

<!-- * Different (sub)disciplines have approached the topic in disparate ways [@riedigerEmotionRegulationAdolescence2014] -->
<!-- * No consistent terminology [@bariolaChildAdolescentEmotion2011] -->
<!-- * No conceptual frameworks [@stifterEmotionRegulation2019] -->
<!-- * No overarching theoretical framework [@bussTheoriesEmotionalDevelopment2019] -->

<!-- Knowledge must be consolidated into overarching theory -->

## Limitations of relevant theory

* Few explicitly address adolescence
* Few comprehensively address predictors of emotion regulation
* None directly guide contemporary research 
* Theories vary widely in scope: Some are broad and non-specific; others describe a specific phenomenon in detail, but lack a broader perspective.
    + Broad theories can frame any research, specific theories generate hypotheses.
    + It would be beneficial to bridge these levels of analysis.

## Towards integrative theory

* First step: identifying relevant phenomena [@borsboomTheoryConstructionMethodology2020]
    + Often done by systematic review
* Narrative reviews: small samples, confirmation bias, emphasize positive results [@littellEvidencebasedBiasedQuality2008]
* Introducing text mining systematic review (TMSR) [@vanlissaMappingPhenomenaRelevant2021]:
    + Unlimited sample size, transparent, objective, reproducible

6653 papers on **emotion regulation** in **adolescence** [age 10-24]

<https://doi.org/10.1007/s40894-021-00160-7>

## Open science

All code and data available at <https://github.com/cjvanlissa/veni_sysrev>

Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]

## Baseline

Manual selection of key constructs from relevant theory and prior reviews [ @bariolaChildAdolescentEmotion2011; @coe-odessEmergentEmotionsAdolescence2019]:

* Bioecological model [@bronfenbrennerBioecologicalModelHuman2007]
* Transactional model [@sameroffUnifiedTheoryDevelopment2010]
* Hall's notion of "storm and stress"
* Theory of normative emotional development [@sroufeEmotionalDevelopmentOrganization1995]
* Tripartite model [@morrisRoleFamilyContext2007]
* Internalization model [@holodynskiDevelopmentEmotionsEmotion2006]
* Polyvagal theory [@porgesOrientingDefensiveWorld1995].
* Model of social-affective engagement and goal flexibility [@croneUnderstandingAdolescencePeriod2012]
* Process model of emotion regulation [@grossHandbookEmotionRegulation2013]
* Social Information Processing Theory [@lemeriseIntegratedModelEmotion2000]

@@

```{r figbaseline, warning=FALSE, eval = TRUE, results='asis', out.width = "100%", fig.cap="Phenomena relevant to adolescents' emotion regulation according to theory (a) and narrative reviews (b; transparent nodes indicate constructs also present in the theory)."}
knitr::include_graphics("presentation/baseline_network.png") 
```

@@

## {.fullslide}

![](presentation/varimps_small.png)

@@

```{r networks1, eval = TRUE, results='asis', out.width = "100%"}
knitr::include_graphics("presentation/tmnetworks_small.png")
```

@@

## Results

* Both analyses reflected some constructs from theoretical literature
    + Particularly related to neurodevelopment and socialization
* Mental health-related outcomes feature prominently
    + Emotion regulation implicated in mental health problems [@leeAdolescentMentalHealth2014],
    + This underlines the importance of research in this area
* Substantial correspondence between keywords and abstracts, suggesting validity of method
* Networks are sparse; few connections among constructs: fragmented literature

@@

## Implications

* Empirical research relies on theory; under-theorized phenomena may be overlooked
    + TMSR offers guidance for phenomena to consider as confounders/causes
* Unembedded terms indicate promising areas of future research
* TMSR network can serve as proto-theoretical nomological network

# Example: Predictors of emotion regulation development

## Person-centered models

```{r, out.height="50%"}
knitr::include_graphics("groups.png")
```

* Predictors of latent growth model?
  + Only linear differences, few predictors (power)
* Multigroup latent growth model?
  + Non-linear differences, but only one moderator
* Latent class growth model with auxilliary variables?
  + Groups with regard to trajectory, not predictors

## Method

SEM-forest [@brandmaierTheoryguidedExplorationStructural2016]

1. Bootstrap sample
2. On each sample, estimate a SEM-model
3. Consider $k$ candidate predictors
4. Identify predictor and value that maximizes $LR$ of post-split multi-group model
5. Average predictions across bootstrap samples

We used 1000 bootstrap samples, $k = 9$

## SEM Forest

```{r, out.width="70%"}
knitr::include_graphics("tree.png")
```

<!-- TS: Confirmatory versus exploratory research: Define the distinction. -->
<!-- TS: Some longitudinal research falls between these categories. For example, complex longitudinal studies, where prior research is sufficient to lead us to hypothesize *some* associations between two constructs, but not sufficiently specific to let us hypothesize which parameters will be significant. -->

<!-- ## Results -->

<!-- * Both analyses reflected some constructs from theoretical literature -->
<!--     + neurodevelopment, socialization -->
<!-- * Undertheorized themes -->
<!--     + Developmental disorders -->
<!--     + Physical health (sic) -->
<!--     + External stressors -->
<!--     + Structural disadvantage -->
<!--     + Addictive behavior -->
<!--     + Identity and moral development -->
<!--     + Sexual development -->

<!-- * This offers guidance for phenomena to consider as relevant predictors -->

## Study design

RQ1: What are the most important predictors of adolescents' trajectories of emotion regulation development?

RQ2: What is the nature of the association of the predictors with the trajectories?

**Participants**

* RADAR data (https://www.uu.nl/en/research/radar)
* 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46)
* Collected between 2006-2011
* Most families were  medium- to high-SES (10% low-SES)

## Candidate predictors

Predictors: 87

Categories:

* Demographics (e.g., SES)
* Biological (age, sex, pubertal development)
* Individual differences (big5, BIS/BAS, empathy)
* Risk behavior (in/externalizing symptoms, substance use, delinquency, school performance, victimization)
* Relationship quality (parents/peers)
* Parenting (perceived and parent-reported)
* Conflict resolution styles

## Dependent variable

Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]

* 24 items (omitted 'emotional awareness' subscale)
* E.g.: “When I’m upset, I have difficulty thinking about anything else”
* 1 (“Almost never”) to 5 (“Almost always”)
* Excellent reliability, $(.94 > \alpha > .95)$.
* Metric invariance across waves
* Quadratic growth model had good fit
  + CFI = .996, TLI = .993, RMSEA = .047, SRMR = 0.020

## Open science

Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]

All code and synthetic data available at

<https://github.com/cjvanlissa/veni_forest>

# Results

## Variable importance

```{r, out.width="90%"}
knitr::include_graphics("pcomb2.png")
```

## Variable importance 2

```{r, out.width="100%"}
knitr::include_graphics("violin.png")
```

## Marginal association

```{r, out.width="40%"}
knitr::include_graphics("plots/pdp1-12.png")
```

## Summary

Important:

* Best predictors: Personality and related constructs
* Best parental predictor is autonomy support (bal. rel.)
  + Negative/intrusive parenting more predictive than other positive practices
* Conflict frequency and behavior
* Empathy, particularly personal distress

@@

## Reflection content

* Proximal factors more predictive of ER development than distal predictors
    + Order mirrors bioecological model
* Most important predictors are routinely assessed (big 5) or overt (conflict)
  + Prime candidates for early risk assessment
  + Conflict resolution behavior can be taught
    - Target for intervention if association is causal
* Emphasis on parenting (in literature and theory) might not be justified
  + Congruent with RI-CLPM showing few parenting effects in adolescence [@vanlissaRoleFathersMothers2019]

## Reflections form

* Many predictors show non-linear effects: Emotion dysregulation only for +/-1SD
* Some predictors show almost no marginal effect (e.g., father's age, drug use)
  + This suggests they might be important in **interactions**

# Limitations and future directions

## Limitations

1. Theories are still verbal and flexible
    * Even if ML leads to new insights, it is not clear how theories should be updated
2. Ignoring causality limits interpretability of machine learning
    * Incorrect direction of causality: meaningless results
    * Conditioning on a collider: biased results

## Future direction 1: Formal theoretical object

Formal theories are expressed in mathematical terms

* unambiguous
* logically sound
* explicates assumptions
* reveals ambiguities and gaps in knowledge
* facilitates conveying ideas to others
* applying rules of formal logic implies testable hypotheses
* refutation of hypotheses dictates specific modifications

## Characteristics of formal theory

According to Meehl:

1. phenomena to be explained
2. putative causal connections between them
3. direction of the effect
4. relative importance of predictors
5. interaction effects
6. functional form of associations

## Formal theory

* Characteristics 1-2 can be described by a causal network (DAG)
* Characteristics 3-6 refer to the nature of associations within that network

Augmented causal networks (ACN, see also Crielaard et al., 2022)

* encode both causal relationships and the nature of those relationships
* imply testable hypotheses
* shared in plain text, e.g., YAML or RDF
* version controlled
* shared conform FAIR software principles,
to be used and updated through distributed collaboration

## Learning a causal network

* Using TMSR, based on what previous authors have written
* Using causal discovery, based on large panel studies
* Using ESM data to leverage precedence in time
    + Causality requires association, precedence in time, and nonspuriousness

## Future direction 2: Functional form

Previous work: Post-hoc interpretation of marginal associations

Future direction: Use a-priori interpretable machine learning

* Symbolic regression applies deep learning or genetic algorithms to learn a mathematical formula that describes the relationship between an outcome and its predictors

>>>>

---
title: "Machine Learning can Advance Theory Formation in the Social Sciences"
subtitle: "Funded by NWO Veni grant VI.Veni.191G.090<br>[doi:10.1002/icd.2370](https://doi.org/10.1002/icd.2370)"
author: "Caspar J. Van Lissa"
institute: "Tilburg University, dept. Methodology & Statistics"
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "docs") })
format: revealjs
bibliography      : ["manuscript/references.bib", "presentation/all_refs.bib"]
---

```{r}
#| label = "setup",
#| include = FALSE
#css: https://maxcdn.bootstrapcdn.com/bootstrap/3.3.7/css/bootstrap.min.css
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

## Central thesis {.smaller}

@@

## Defining theory {.smaller}

@@

```{r}
#| out.width = "40%"
knitr::include_graphics("presentation/empiricalcycle.png")
```

* Empirical cycle [@degrootMethodologieGrondslagenVan1961]
* Lack of theory cannot be overcome by improving deductive practices
* Theory formation requires inductive (exploratory) research [@creswellDesigningConductingMixed2017]

@@

**NEW**: Epistemological implications of ML for theory formation [@vanlissaDevelopmentalDataScience2022]

@@

## Holistic approach {.smaller}

@@

* Good theory incorporates most important causes
* Cannot assess the relative utility of predictors across studies
* Include potential predictors from various theories, and undertheorized factors
* Pitfalls:
    + Multicollinearity
    + Colliders

@@

## Person-centered approaches {.smaller}

@@

* E.g., latent class analyses
    + "Which individuals are similar?"
    + Other *unsupervised learning* methods exist
* Tree-based models
    + "Why are these individuals similar?"
    + Group individuals based on predictors to maximize homogeneity of the outcome
* RI-CLPM and DSEM
    + Rarely explain heterogeneity in within-person effects
    + Regularization

@@

When using data models to guide theory formation, generalizability is essential [@haslbeckModelingPsychopathologyData2021]

@@

* cross-validation to estimate predictive accuracy [@hastieElementsStatisticalLearning2009]

@@

```{r}
#| out.width = "40%"
library(ggplot2)
colors <- c("Variance" = "red", "Bias" = "green", "Total error" = "black")

@@

## From models to theory {.smaller}

@@

* Abductive formal theory construction (AFTC) framework
    + construct formal theory
    + Generate data
    + Mathematically compare to empirical data
    + like a posterior predictive check in Bayes
    + Discrepancies: amend formal theory

@@

## Text Mining Systematic Review

* Identify which constructs are discussed in 6653 papers on **emotion regulation** in **adolescence** [age 10-24]
* Identify which constructs co-occur in papers
* Compare resulting networks to construct networks from existing theories and systematic reviews
* Paper: <https://doi.org/10.1007/s40894-021-00160-7>
* Code and data: <https://github.com/cjvanlissa/veni_sysrev>
* Workflow for Open Reproducible Code in Science (WORCS) used to make analyses reproducible [@vanlissaWORCSWorkflowOpen2020]

@@

```{r}
#| label = "figbaseline",
#| warning = FALSE,
#| eval = TRUE,
#| results = "asis",
#| out.width = "100%",
#| fig.cap = "Phenomena relevant to adolescents' emotion regulation according to theory
#| (a) and narrative reviews (b; transparent nodes indicate constructs also present in
#| the theory)."
knitr::include_graphics("presentation/baseline_network.png") 
```

@@

@@

```{r}
#| label = "networks1",
#| eval = TRUE,
#| results = "asis",
#| out.width = "100%"
knitr::include_graphics("presentation/tmnetworks_small.png")
```

@@

## Results {.smaller}

* Both analyses reflected constructs from theoretical literature
    + Especially pertaining to neurodevelopment and socialization
* Mental health-related outcomes feature prominently
    + Emotion regulation implicated in mental health problems [@leeAdolescentMentalHealth2014],
    + This underlines the importance of research in this area
* Substantial correspondence between keywords and abstracts, suggesting validity of method
* Networks are sparse; few connections among constructs: fragmented literature

@@

<!-- ## Implications -->

<!-- * Empirical research relies on theory; under-theorized phenomena may be overlooked -->
<!--     + TMSR offers guidance for phenomena to consider as confounders/causes -->
<!-- * Unembedded terms indicate promising areas of future research -->
<!-- * TMSR network can serve as proto-theoretical nomological network -->

## Predicting ER development {.smaller}

* What are important predictors of adolescents' ER development?
* What is the nature of the association of the predictors with the trajectories?
* 497 Dutch adolescents (283 boys; age at T1: M =13.03, SD = 0.46) 
* Predictors: 87 (Demographics, Biological, Individual differences, Risk behavior, Relationship quality, Parenting, Conflict resolution styles)
* Outcome: Difficulties in emotion regulation [@gratzMultidimensionalAssessmentEmotion2004]
* Code and synthetic data: <https://github.com/cjvanlissa/veni_forest>

## SEM Forests

```{r}
#| out.width = "70%"
knitr::include_graphics("tree.png")
```

## Results {.smaller}

* Best predictors: Personality and conflict resolution behavior
* Negative parenting more predictive than other positive practices

@@

* Many predictors show non-linear effects

## Conclusions

* Machine learning reveals blind spots in existing theory
    + Missing constructs, relative importance, functional form
* ML can detect patterns in published research and primary data
* "Inductive methods" for consolidating insights into theory remain vague
    + Theories are still verbal and flexible, how should they be updated?
* Ignoring causality limits interpretability of machine learning
    + Incorrect direction of causality: meaningless results
    + Conditioning on a collider: biased results

<!-- ## Future direction 1: Formal theoretical object -->

<!-- Formal theories are expressed in mathematical terms -->

<!-- * unambiguous -->
<!-- * logically sound -->
<!-- * explicates assumptions -->
<!-- * reveals ambiguities and gaps in knowledge -->
<!-- * facilitates conveying ideas to others -->
<!-- * applying rules of formal logic implies testable hypotheses -->
<!-- * refutation of hypotheses dictates specific modifications -->

<!-- ## Characteristics of formal theory -->

<!-- According to Meehl: -->

<!-- 1. phenomena to be explained -->
<!-- 2. putative causal connections between them -->
<!-- 3. direction of the effect -->
<!-- 4. relative importance of predictors -->
<!-- 5. interaction effects -->
<!-- 6. functional form of associations -->

<!-- ## Formal theory -->

<!-- * Characteristics 1-2 can be described by a causal network (DAG) -->
<!-- * Characteristics 3-6 refer to the nature of associations within that network -->

<!-- Augmented causal networks (ACN, see also Crielaard et al., 2022) -->

<!-- * encode both causal relationships and the nature of those relationships -->
<!-- * imply testable hypotheses -->
<!-- * shared in plain text, e.g., YAML or RDF -->
<!-- * version controlled -->
<!-- * shared conform FAIR software principles, -->
<!-- to be used and updated through distributed collaboration -->

<!-- ## Learning a causal network -->

<!-- * Using TMSR, based on what previous authors have written -->
<!-- * Using causal discovery, based on large panel studies -->
<!-- * Using ESM data to leverage precedence in time -->
<!--     + Causality requires association, precedence in time, and nonspuriousness -->

<!-- ## Future direction 2: Functional form -->

<!-- Previous work: Post-hoc interpretation of marginal associations -->

<!-- Future direction: Use a-priori interpretable machine learning -->

<!-- * Symbolic regression applies deep learning or genetic algorithms to learn a mathematical formula that describes the relationship between an outcome and its predictors -->
